Modelling at-bats in baseball using the generalised Pareto distribution

### 2023-06-02 bibat/examples/baseball/docs/report.qmd
Fix include in baseball report
update baseball example

---
title: "Modelling at-bats in baseball using the generalised Pareto distribution"
subtitle: "A report created using bibat version 0.1.3"
author: "Teddy Groves"
bibliography: bibliography.bib
reference-location: margin
toc: true
format:
  html:
    embed-resources: true
filters:
  - include-code-files
---

@@

Still, there were a few problems. In particular the data were typically
unbalanced because better players tended to fproduce more data than worse
players. The result of this was that my models would often inappropriately think
the bad players were like the good players: they would not only tend to be too
certain about the abilities of low-data players, but also be biased, thinking
that these players are probably a bit better than they actually are. I never
came up with a good way to solve this problem, despite trying a lot of things!

@@

```{.zsh}
baseball
├── CODE_OF_CONDUCT.md
├── LICENSE
├── Makefile
├── README.md
├── baseball
│   ├── __init__.py
│   ├── data_preparation.py
│   ├── inference_configuration.py
│   ├── investigate.ipynb
│   ├── notebooks
│   ├── prepare_data.py
│   ├── readme.md
│   ├── sample.py
│   ├── stan
│   │   ├── custom_functions.stan
│   │   ├── multilevel-linear-regression.stan
│   │   └── readme.md
│   ├── stan_input_functions.py
│   └── util.py
├── bibat_version.txt
├── data
│   └── raw
│       ├── raw_measurements.csv
│       └── readme.md
├── docs
│   ├── bibliography.bib
│   ├── img
│   │   ├── example.png
│   │   └── readme.md
│   └── report.qmd
├── inferences
│   ├── fake_interaction
│   │   └── config.toml
│   ├── interaction
│   │   └── config.toml
│   └── no_interaction
│       └── config.toml
├── plots
│   ├── posterior_ll_comparison.png
│   └── posterior_predictive_comparison.png
├── pyproject.toml
└── requirements.txt
```

This folder implements bibat's example analysis - a comparison of linear
regression with two different design matrices. To check that everything was
working correctly I ran the analysis like this:

@@

Finally I removed the example analysis's raw data:

@@

```{.python include="../baseball/data_preparation.py" start-line=35 end-line=53}
```

@@

To check that all this worked, I ran the data preparation script manually:^[I
could also have just run `make analysis` again]

@@

I wanted to test two statistical models: one with the modelling the distribution
of per-player logit-scale at-bat success rates as a normal distbution with
unknown mean and standard deviation, and another where the same logit-scale
rates have a generalised pareto distribution.

So, genve a table of $N$ player profiles, with each player has $y$ successes out
of $K$ at-bats and an unknown latent success rate $\alpha$, I wanted to use this
measurement model:

@@

In the generalised pareto model I would give the $\alpha$s this prior model,
with the hyperparameter $\text{min }\alpha$ assumed to be known exactly and $k$
and $\sigma$ given prior distributions that put the $\alpha$s in the generally
plausible range of between roughly 0.1 and 0.4.

@@

I started by defining some Stan input functions that pass arbitary prepared
data on to each of the models:^[Note that this code uses the scipy function
`logit`, which it imported like this: `from scipy.special import logit`]

@@

But why stop there? It can also be useful to generate Stan input data from a
model with a model, by running it in simulation model with hardcoded parameter
values. Here are some functions that do this for both of our models:

@@

Now all the building blocks for making statistical inferences - raw data, data
preparation rules, statistical models and recipes for turning prepared data
into model inputs - were in place. The last step before actually running Stan
was to write down how put the blocks together. Bibat has another concept for
this, called 'inferences'.

@@

- The Stan file, prepared data folder and stan input function are referred to by
  strings. The analysis should raise an error if you enter a non-existing value.
- Both inferences are set to run in "prior" and "posterior" modes - the other
  pre-defined mode is "kfold", but you can also write your own!
- You can enter arbitrary arguments to cmdstanpy's `CmdStanModel.sample` method
  in the `[sample_kwargs]` table.
- You can enter mode-specific overrides in `[sample_kwargs.<MODE>]`. This can be
  handy if you want to run more or fewer iterations for a certain mode.

Now when I ran `make analysis` again, I saw messages indicating that Stan had
run, and found that the `inferences` subfolders had been populated:

@@

The final step was to document my analysis. To do this I edited the file
`docs/report.qmd`, then ran `quarto render docs/report.qmd`, which produced the
very html document that you are probably reading now! You can find the complete
`report.qmd` file
[here](https://github.com/teddygroves/bibat/blob/main/bibat/examples/baseball/docs/report.qmd).

>>>>

---
title: "Modelling at-bats in baseball using the generalised Pareto distribution"
subtitle: "A report created using bibat version 0.1.6"
author: "Teddy Groves"
bibliography: bibliography.bib
reference-location: margin
toc: true
format:
  html:
    embed-resources: true
filters:
  - include-code-files
---

@@

Still, there were a few problems. In particular the data were typically
unbalanced because better players tended to produce more data than worse
players. The result of this was that my models would often inappropriately think
the bad players were like the good players: they would not only tend to be too
certain about the abilities of low-data players, but also be biased, thinking
that these players are probably a bit better than they actually are. I never
came up with a good way to solve this problem, despite trying a lot of things!

@@

```{.zsh}
baseball
├── CODE_OF_CONDUCT.md
├── LICENSE
├── Makefile
├── README.md
├── bibat_version.txt
├── data
│   └── raw
│       ├── raw_measurements.csv
│       └── readme.md
├── docs
│   ├── bibliography.bib
│   ├── img
│   │   ├── example.png
│   │   └── readme.md
│   ├── report.html
│   ├── report.pdf
│   └── report.qmd
├── baseball
│   ├── __init__.py
│   ├── data_preparation.py
│   ├── fitting_mode.py
│   ├── inference_configuration.py
│   ├── investigate.ipynb
│   ├── notebooks
│   ├── prepare_data.py
│   ├── readme.md
│   ├── sample.py
│   ├── stan
│   │   ├── custom_functions.stan
│   │   ├── multilevel-linear-regression.stan
│   │   └── readme.md
│   ├── stan_input_functions.py
│   └── util.py
├── inferences
│   ├── fake_interaction
│   │   └── config.toml
│   ├── interaction
│   │   └── config.toml
│   └── no_interaction
│       └── config.toml
├── plots
│   ├── posterior_ll_comparison.png
│   └── posterior_predictive_comparison.png
└── pyproject.toml
```

This folder implements bibat's example analysis – a comparison of linear
regression with two different design matrices. To check that everything was
working correctly I ran the analysis like this:

@@

Finally, I removed the example analysis's raw data:

@@

```{.python include="../baseball/data_preparation.py" start-line=35 end-line=54}
```

@@

To check that all this worked, I ran the data preparation script manually:^[I
could also have just run `make analysis` again. This would have caused an error
on the step after prepare_data.py, which is fine!]

@@

I wanted to test two statistical models: one with the modelling the distribution
of per-player logit-scale at-bat success rates as a normal distribution with
unknown mean and standard deviation, and another where the same logit-scale
rates have a generalised Pareto distribution.

So, given a table of $N$ player profiles, with each player has $y$ successes out
of $K$ at-bats and an unknown latent success rate $\alpha$, I wanted to use this
measurement model:

@@

In the generalised Pareto model I would give the $\alpha$s this prior model,
with the hyperparameter $\text{min }\alpha$ assumed to be known exactly and $k$
and $\sigma$ given prior distributions that put the $\alpha$s in the generally
plausible range of between roughly 0.1 and 0.4.

@@

I started by defining some Stan input functions that pass arbitrary prepared
data on to each of the models:^[Note that this code uses the scipy function
`logit`, which it imported like this: `from scipy.special import logit`]

@@

But why stop there? It can also be useful to generate Stan input data with a
model, by running it in simulation mode with hardcoded parameter values. Here
are some functions that do this for both of our models:

@@

Now all the building blocks for making statistical inferences – raw data, data
preparation rules, statistical models and recipes for turning prepared data
into model inputs – were in place. The last step before actually running Stan
was to write down how put the blocks together. Bibat has another concept for
this, called 'inferences'.

@@

- The Stan file, prepared data folder and Stan input function are referred to by
  strings. The analysis should raise an error if you enter a non-existing value.
- Both inferences are set to run in "prior" and "posterior" modes – the other
  pre-defined mode is "kfold", but you can also write your own!
- You can enter arbitrary arguments to cmdstanpy's `CmdStanModel.sample` method
  in the `[sample_kwargs]` table.
- You can enter mode-specific overrides in `[mode_options.<MODE>]`. This can be
  handy if you want to run more or fewer iterations for a certain mode.

With all these changes made I ran `make analysis` again. There were no errors
until after sampling, which was expected as I hadn't yet customised the
investigation code, and I saw messages indicating that Stan had run. I also
found that the `inferences` subfolders had been populated:

@@

The final step was to document my analysis. To do this I edited the file
`docs/report.qmd`, then ran `quarto render docs/report.qmd`, which produced the
very HTML document that you are probably reading now! You can find the complete
`report.qmd` file
[here](https://github.com/teddygroves/bibat/blob/main/bibat/examples/baseball/docs/report.qmd).
